Road Segment Prioritization for Bicycle Infrastructure

### 2021-01-19 README.Rmd
updated paper based on comments

---
title: Road Segment Prioritization for Bicycle Infrastructure
date: "`r Sys.Date()`"
geometry: margin=2cm
bibliography:
  - mybibfile.bib
  - rlbibfile.bib
#linenumbers: true
numbersections: true
abstract: "Understanding the motivators and deterrents to cycling is essential for creating infrastructure that gets more people to adopt cycling as a mode of transport. The aim of this paper is to develop methods to support the prioritisation of cycling infrastructure and cycling network design, accounting for cyclist preferences and the growing emphasis on 'filtered permeability' and 'Low Traffic Neighborhood' interventions internationally. Distance decay, routing, and network analysis techniques are used to examine where future cycling demand is most likely to arise, how such demand could be accomodated within existing street networks, and how to ensure a fair distribution of investment. The city of Manchester is used as a case study to demonstrate the methods which, as demonstrated in reproducible code and open access data accompanying the paper, can be applied to any city. We find that flexible routing profiles that are adjustable to local needs can avoid high-stress road segments, enabling the identification of networks that balance directness against the need for safe and stress-free routes. Community detection is used to partition the city and evaluate results from 'utilitarian' and 'egalitarian' policy perspectives. Based on discussion of the findings from a policy perspective, we conclude that the 'egalitarian' model prioritisation approach demonstrated in this paper has great potential to support cost-effective strategic cycle network planning internationally. \\par\\textbf{Keywords:} cycling networks, low-traffic neighborhoods, routing, transport equity"
# keywords: "cycling, low-traffic neighborhoods, potential demand, routing"
output:
  # bookdown::github_document2:
  bookdown::pdf_document2:
  # # bookdown::word_document2:
      toc: false
      extra_dependencies: ["subcaption", "longtable", "float", "multirow", "placeins", "multicol"]
---

@@

- Intro: possible trim down substantially then add some literature in (a) Calculating Potential Demand [DONE]/ (b) Routing [DONE]/ (c) Road Segment Prioritization
- Community Detection: Where in the document should this section be? It is now before Road Segment Prioritization, but I can merge them
- Should we use other cities for the sake of comparison (compare communities, road types etc)
- How to add tables properly (they are now added using latex code which only shows up when outputting to pdf)
- Footnotes are sometimes split between pages. How do I keep them on the same page?
- What is the best way to write pseudocode (Steps for Algorithm 1 and 2)

@@

In this context, interest and investment in active modes is growing. The benefits of the latter extend beyond congestion and the environment, as it promises to help alleviate what is referred to as the pandemic of global inactivity; physical inactivity is on the rise and has become the 4th highest cause of death globally [@kohl2012pandemic]. Various studies have documented the association between active transport and lower risk of disease, including cancer and cardiovascular disease [@celis2017association; @jarrett2012effect; @patterson2020associations].
In the wake of the Covid-19 pandemic, and the resulting reduced capacity of public transport, the UK government has pledged to invest billions of pounds to improve walking and cycling infrastructure across the country. While this unprecedented sum is an opportunity to reshape cities in a way that improves the well-being of citizens, it does come with a warning:

@@

A considerable amount of research has been done towards that end. Segregated cycling infrastructure^[*Segregated cycling infrastructure* refers to road space that is allocated to cyclists only, with physical separation to protect cyclists from other modes of transport.] has been shown to increase cycling uptake [@aldred2019impacts; @goodman2014new; @marques2015infrastructure], with the separation from motorized vehicles being key [@winters2011motivators]. Revealed preference of cyclists shows that they are willing to deviate from the most efficient routes in order to commute on safer roads [@crane2017longitudinal]. However, such deviations are only considered if they do not considerably increase route circuitry; behaviour studies have found that the probability of choosing a route decreases in proportion to its length relative to the shortest route [@broach2011bicycle; @winters2010far]. Another defining feature for cycling infrastructure is how well connected it is. Cyclists prefer cohesive infrastructure, particularly when cycling on arterial roads with high levels of motorized traffic [@stinson2003commuter], and the lack of well-connected cycling infrastructure is one of the main obstacles to increasing cycling uptake [@caulfield2012determining]. While direct and cohesive cycling networks have been shown to positively impact cycling rates, density^[making an area's bicycle network denser means adding more cycling routes in the area and thereby giving cyclists more route options] of the cycling network is also vital [@schoner2014missing].

As noted by @buehler2016bikeway, there has been a shift in focus from localized (street or intersection) level planning towards planning based on studying the cycling network as a whole since the turn of the 21^st^ century. They emphasize the promise of this shift in capturing the network-wide effect of street-level interventions. *Optimization* techniques have been used to propose improvements to cycling networks. @mesbah2012bilevel propose a bi-level formulation to optimize allocation of cycling lanes to the network without exceeding a set budget. The upper level is the proposed interventions and the lower level is the route choices made by users in reaction to changes in the network. The problem accounts for the effect of cycling lanes on car traffic, and attempts to maximize utilization of said lanes with minimal impact on car travel times. To improve cohesion of the suggested network, a constraint is added so that each link^[*link* refers to a road segment throughout this research] with a bike lane should be connected to at least one destination. Car usage is not considered by @mauttone2017bicycle, who develop an optimization framework that aims to minimize the total user cost of cycling on the network. The aggregate flow on links is obtained by using shortest paths to route existing cycling demand onto the road network, and the solution is a proposed set of links where cycling infrastructure should be added in order to minimize the overall travel cost of cyclists across the network. The cost of traversing a link is given as a function of its length and whether or not it has cycling infrastructure, and a discontinuity penalty is also added to prioritize connected road segments. The problem has also been solved by attempting to find the minimum cost of improving roadway links to meet a desired level of service (LOS) [@duthie2014optimization]. In this formulation, all OD pairs need to be connected by roads that meet the desired LOS, and a directness constraint is added so that paths between OD pairs do not exceed a certain multiple of the shortest path.

These problem formulations do not explicitly solve for continuity, which is dealt with using a either (a) a constraint specifying that each link with a bike lane should be connected to at least one destination [@mesbah2012bilevel], (b) a constraint on deviation from shortest paths [@duthie2014optimization], or (c) a discontinuity penalty [@mauttone2017bicycle]. To solve for continuity, the graph-theoretic concept of *connected components*, has been used. @natera2019data study the existing cycling network in terms of its disconnected components and introduce two different algorithms to connect these components by their most critical links and, in doing so, measure the size of the growth of the largest connected component as a function of the kilometers of network added. They observe that small investments at strategic points have a large impact on connectivity in most cases. The concept of connected components is also at the core of the methodology proposed by @olmos2020data. After routing the cycling demand onto the network links, they use percolation theory to filter out the links based on the aggregate flow^[*flow* is used throughout this research to refer to the cycling demand when it is routed onto the road network. The flow on any road segment is the cumulative demand on it, resulting from cyclists commuting between various OD pairs] passing through them, varying the flow threshold for filtering to identify the minimum flow at which the whole city is connected by a giant component. The results show a cycling network that connects the entire city, and subtracting links intersecting with current cycling infrastructure identifies links proposed for intervention.

The problem formulations outlined above look at the network as a whole when attempting to improve it. An alternative approach is to identify the different sub-networks that exist within the larger network, and work on improving each separately. Trip patterns in a city are not uniformly distributed geographically, and *community finding* methods have been used to partition study areas into localized areas that experience a disproportionate number of trips within them. @akbarzadeh2018designing use a modularity maximization approach [@blondel2008fast] on taxi trip data to identify 7 different communities in the city of Isfahan, Iran. An optimization problem is then formulated to connect nodes within each community with cycling infrastructure. The emphasis is on connectivity within the communities, not between them. @bao2017planning adopt a similar methodology, but use hierarchical clustering to specify the desired number of clusters. They use a greedy network expansion algorithm, where the link with the highest benefit-cost ratio in each cluster is selected, and the network is grown by adding neighboring links to the solution until a budget limit is met. The benefit is the flow on the link, and each link is assigned a cost based on current road conditions.

All of these network-level methodologies are underpinned by different ethical principles, even though these principles are not explicitly acknowledged by the authors. This is important since different ethical principles constitute different problem formulations and targets. Broadly speaking, transport appraisal can be based on either utilitarian or egalitarian principles. The former seeks to maximize the overall benefit, while the latter is concerned with a fair distribution of benefits [@jafino2020transport]. @nahmias2017integrating criticize the utilitarian approach that has been historically popular in the evaluation of transport investments, explaining how the maximization of overall benefit fails to account for the distribution of that benefit among communities or individuals. @lucas2016method explain how transport studies have traditionally looked at the bigger picture without studying the distribution of investments on the different parts of the study area, and go on to propose an egalitarian approach that ensures the dis-aggregation of transport policy benefits across the study area. @pereira2017distributive also emphasize the need for a more egalitarian approach to transport planning. They highlight accessibility as a cornerstone of distributive justice, and contend that policies should aim to distribute investments in a way that minimizes spatial variations in accessibility. This research attempts to compare the two principles, and in doing so determine whether a methodology formulated based on an egalitarian approach can be feasible in designing a cycling network that aligns with motivators and deterrents to cycling.

<!-- # Data and Geographical Scale of Analysis -->

The approach is based on origin-destination (OD), which can be obtained from many sources.
In this paper we use open access data from the UK census [@ONS2011flowdata], which contains aggregate statistics on number of commuters between Middle layer Super Output Area (MSOA) zones, which have an average population of just over 8000 [@ofn2018population].

@@

Relying only on existing cycling activity to prioritize cycling infrastructure can reinforce existing cycling patterns and ignores potential cycling demand that could be satisfied by a connected network. To avoid this issue, @duthie2014optimization choose to ignore existing demand completely, and focus on creating a network that connects the entire study area. @olmos2020data obtain the distance distribution of cyclists using a smartphone-based bicycle GPS data, and then use a rejection-sampling algorithm on the OD data of the study area to match the potential demand distribution to the distribution obtained from GPS data.

For our purposes, we use a logistic regression model to calculate potential cycling demand. The model was adopted from the Propensity to Cycle Tool (PCT), which estimates the proportion of trips ($\boldsymbol{C_{p}}$) for each origin-destination pair that would cycle under different scenarios of change (we used the Government Target scenario, representing a nationwide target of doubling cycling by 2025) as a function of distance and hilliness [@lovelace2017propensity]. The logistic regression model used to calculate $\boldsymbol{C_{p}}$ has the following parameters:

@@

It should be noted that the calculations assume a future that is constrained by physical geography; i.e. we consider cycling in the traditional sense. Recently there have been various micro-mobility solutions, including e-bikes, that allow commuters to traverse longer distances and hillier roads with less effort than traditional bicycles. While these modes would probably be associated with less geographical impedance, it is beyond the scope of this work to integrate that into the analysis. Doing so is partially restricted by the lack of data on the proliferation of these modes, which raises the point that perhaps the census data category of `Bicycle' is too vague, and should be further dis-aggregated to distinguish between traditional bicycles and other forms of micro-mobility.

@@

The next step in our approach is to route the potential cycling demand ($\boldsymbol{C_{p}}$) between all OD pairs onto the road network. We choose not to use the PCT approach for routing as it is reliant on external and therefore inflexible routing services.
<!-- This expands on the work of @mauttone2017bicycle, by going beyond simply favoring roads with existing cycling infrastructure to creating a hierarchy of road preference. -->

@@

\noindent A weighted distance $\boldsymbol{d_{w}}$ for each road segment is calculated as following ^[The **dodgr** r package [@padgham2019dodgr] is used to route cycling demand onto the road network. The package uses the OpenStreetMaps (OSM) road network and allows the user to assign weights to roads based on their type. The routing is done based on weighted shortest paths, with the distance along each road segment being divided by a factor to obtain the weighted distance for routing. It is more intuitive to multiply when weighting a network, but the dodgr package divides by numbers between 0 and 1, which achieves the same result. For the sake of reproducibility, we stick to the convention used in the package.]:

@@

```{r flowsfacetunweighted, fig.cap="Flow Results Based on **Unweighted** Shortest Paths (Manchester)", out.width="90%", fig.show='hold', echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/flows_facet_unweighted_Manchester.png")
```

```{r flowsfacetweighted, fig.cap="Flow Results Based on **Weighted** Shortest Paths (Manchester)", out.width="90%", fig.show='hold', echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/flows_facet_weighted_Manchester.png")
```

@@

A major challenge facing ‘top-down’ planning approaches is how to incorporate egalitarian principles by fairly distributing investments in cycling infrastructure. One way of quantifying this is to split up the city into smaller geospatial areas and target equal investment in each of those areas, , which could also help ensure that on-the-ground surveys are made by local stakeholders, an important component of the planning process  [@parkin2018designing]. Community detection offers us a way to delineate such a split; cyclists are limited in their commuting distance (see Figure \@ref(fig:cyclinghistmanc)), and so trip attractors are more likely to have a local catchment area of cyclists.

@@

In our case, the network is the city; the nodes are the population-weighted MSOA centroids and the links connecting each MSOA pair are weighted by the potential cycling demand between them. The Louvian method  [@blondel2008fast] is used to separate MSOAs into communities. Potential cycling demand is used since we assume that this is what the cycling demand will be once the cycling infrastructure is added. To assign road links to communities, the following steps are carried out:

<!-- 1. Create links between MSOA centroids and weigh these links by potential cycling demand between them.
2. Use Louvian method to determine optimal number of communities and assign each MSOA centroid to a community.
3. Assign each road link to the same community as the closest MSOA centroid to it. -->
```{r, eval = FALSE}
1. Create links between MSOA centroids and weigh these links by potential cycling demand between them.
2. Use Louvian method to determine optimal number of communities and assign each MSOA centroid to a community.
3. Assign each road link to the same community as the closest MSOA centroid to it.
```

@@

```{r growthtotal, fig.cap="Comparing Overall (Dashed) and Community Level Person-Km Satisfied (Manchester)", out.width="45%", fig.show='hold', echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/Growth_Results/growth_utilitarian_satisfied_km_both_flow_column.png",
  "data/Manchester/Plots/Growth_Results/growth_egalitarian_satisfied_km_both_flow_column.png"
))
```

@@

```{r growth3MapandBar, fig.cap="Results of Alg. 2 (Manchester)", out.width="45%", fig.show='hold', echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/Growth_Results/growth_egalitarian_priority_all_FLOW.png",
  "data/Manchester/Plots/Growth_Results/growth_egalitarian_investment_highways_flow.png"
))
```

@@

```{r componentsandGCC, fig.cap="Network Characteristics", out.width="45%", fig.show='hold', echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/Growth_Results/growth_util_egal_components_number_comparisonManchester.png",
  "data/Manchester/Plots/Growth_Results/growth_util_egal_components_gcc_comparisonManchester.png"
))
```

@@

While segregated, connected, and direct cycling infrastructure is key to achieving high levels of cycling, research has shown that it cannot exist in a vacuum. @wardman2007factors developed a mode choice model for the UK and their results showed that improved cycling infrastructure on its own only had modest impacts on mode shift, and even the unlikely scenario of all urban routes being serviced by segregated bike lanes was forecast to increase cycling mode share by only 3%. International research shows that cities that invest in more comprehensive cycling projects show a more significant increase in the number of cyclists as well as the cycling mode share [@parkin2018designing; @pucher2010infrastructure]. These cities do not just focus on infrastructure, but on general policies as well as restricting car use. Evaluation of policies in Denmark and Germany and the Netherlands has shown that their high cycling mode share is down to a broader set of policies that also include traffic calming, filtered permeability interventions, cycling rights of way, bike parking, integration with the public transport network, and making driving cars both expensive and inconvenient [@pucher2008making]. While these policies are outside the scope of this research, it is important to recognize their key role in bringing about an increase in levels of cycling.

@@

This paper provides a methodology for prioritizing road segments for investments in cycling infrastructure, with the output being a proposed cycling network for the city under study. 
The approach aims to respect both the needs of the users and the ambitions of stakeholders working at local/regional levels.
The results, primarily detailed route network maps based on current travel behaviour derived from origin-destination data, can provide an evidence-based starting point for introducing a direct, connected, and low-stress network for any city by leveraging available open data. Given that cycle network development is seen as the “most essential activity entailed in the design of cycle-friendly infrastructure” [@parkin2018designing], we believe that the methods have great potential to inform investment in cities such as Manchester where there is political will to invest in cycling long-term. A benefit of the approach is that it has relatively modest data requirements: only the road network (from OSM), its topography (from satellite imagery), and origin-destination data (from the national census in this case) are needed, opening up the possibility of deploying the approach in other cities^[The results are easily reproducible for all UK cities, and can also be reproduced for cities elsewhere given the availability of commuter data. For instructions on reproducing the results shown in this paper, see the **README** file in the R folder of the .zip file that accompanies this paper (to be released as open source software when the paper is published).].

@@

A core part of the methodology is determining where there is high potential for cycling, and using this as a basis for recommendations on road space reallocation in order to unlock potential cycling demand. A routing engine was used to study the effects of limiting access on different road types, and it was found that reducing cyclist flow on roads with high through-traffic resulted in acceptable increases in commuting distances. As a result, a hierarchy of road preference was used to route potential cycling demand. Algorithms were developed to determine where investments in cycling infrastructure should be prioritized. These algorithms, based on connectivity and egalitarian principles of resource distribution, ensured that whatever the level of investment, the resulting cycling network would improve connectivity of existing cycling infrastructure.

>>>>

---
title: Road Segment Prioritization for Bicycle Infrastructure
date: "`r Sys.Date()`"
geometry: margin=2cm
bibliography:
  - mybibfile.bib
  - rlbibfile.bib
#linenumbers: true
numbersections: true
abstract: "Understanding the motivators and deterrents to cycling is essential for creating infrastructure that gets more people to adopt cycling as a mode of transport. The aim of this paper is to develop methods to support the prioritisation of cycling infrastructure and cycling network design, accounting for cyclist preferences and the growing emphasis on 'filtered permeability' and 'Low Traffic Neighborhood' interventions internationally. Distance decay, routing, and network analysis techniques are used to examine where future cycling demand is most likely to arise, how such demand could be accommodated within existing street networks, and how to ensure a fair distribution of investment. The city of Manchester is used as a case study, although the methods can be applied to any city, as demonstrated in a reproducible code and open access data accompanying the paper. We find that flexible routing profiles that are adjustable to local needs can avoid road segments that are stressful to cycle on, enabling the identification of networks that balance directness against the need for safe and stress-free routes. We assess two different policy interventions , 'utilitarian' and 'egalitarian' by analyzing the connectivity of the city through community detection. From a policy perspective, our findings indicate that the 'egalitarian' prioritisation model approach is the more adequate to support cost-effective strategic cycle network planning. \\par\\textbf{Keywords:} cycling networks, low-traffic neighborhoods, routing, transport equity"
# keywords: "cycling, low-traffic neighborhoods, potential demand, routing"
output:
  # bookdown::github_document2:
  bookdown::pdf_document2:
  # # bookdown::word_document2:
      toc: false
      extra_dependencies: ["subcaption", "longtable", "float", "multirow", "placeins", "multicol"]
---

@@

<!-- - Intro: possible trim down substantially then add some literature in (a) Calculating Potential Demand [DONE]/ (b) Routing [DONE]/ (c) Road Segment Prioritization -->
- Community Detection: Where in the document should this section be? It is now before Road Segment Prioritization, but I can merge them
<!-- - Should we use other cities for the sake of comparison (compare communities, road types etc) -->
- How to add tables properly (they are now added using latex code which only shows up when outputting to pdf)
- Footnotes are sometimes split between pages. How do I keep them on the same page?
<!-- - What is the best way to write pseudocode (Steps for Algorithm 1 and 2) -->

@@

In this context, interest and investment in active modes are growing. The benefits extend beyond congestion and the environment, as active travel can also help alleviate what is referred to as the pandemic of global inactivity; physical inactivity is on the rise and has become the 4th highest cause of death globally [@kohl2012pandemic]. Various studies have documented the association between active transport and lower risk of disease, including cancer and cardiovascular disease [@celis2017association; @jarrett2012effect; @patterson2020associations].
In the wake of the Covid-19 pandemic, and the resulting reduced capacity of public transport, the UK government has pledged to invest billions of pounds to improve walking and cycling infrastructure across the country. While this unprecedented sum is an opportunity to reshape cities in a way that improves the well-being of citizens, it does come with a warning:

@@

## Motivators and Deterrents to Cycling

A considerable amount of research has been done towards that end. Segregated cycling infrastructure^[*Segregated cycling infrastructure* refers to road space that is allocated to cyclists only, with physical separation to protect cyclists from other modes of transport.] has been shown to increase cycling uptake [@aldred2019impacts; @goodman2014new; @marques2015infrastructure], with the separation from motorized vehicles being key [@winters2011motivators]. Revealed preference of cyclists shows that they are willing to deviate from the most efficient routes in order to commute on safer roads [@crane2017longitudinal]. However, such deviations are only considered if they do not considerably increase route circuitry; behaviour studies have found that the probability of choosing a route decreases in proportion to its length relative to the shortest route [@broach2011bicycle; @winters2010far]. Another defining feature for cycling infrastructure is how well connected it is. Cyclists prefer cohesive infrastructure, particularly when cycling on arterial roads with high levels of motorized traffic [@stinson2003commuter]. The lack of well-connected cycling infrastructure is one of the main obstacles to increasing cycling uptake [@caulfield2012determining]. While direct and cohesive cycling networks have been shown to positively impact cycling rates, density^[making an area's bicycle network denser means adding more cycling routes in the area and thereby giving cyclists more route options] of the cycling network is also vital [@schoner2014missing].

## Network-Level Approaches

The studies outlined above lay out the fundamentals for designing cycling networks that generate significant cycling uptake, but they do not propose network-level interventions. In this section we outline techniques used in past studies, namely optimization and network analysis techniques, such as connected components and community detection, and examine how they are leveraged to suggest cycling network designs. We compare the effectiveness of these network-level studies in incorporating the fundamentals outlined above. The methodology of this research is inspired by these techniques, but it attempts to add to them by ensuring that all of the outlined fundamentals are accounted for. It also goes further by attempting to factor in ethical considerations relating to distribution of investment.

*Optimization* techniques have been used to propose improvements to cycling networks. @mesbah2012bilevel propose a bi-level formulation to optimize allocation of cycling lanes to the network without exceeding a set budget. They account for the effect of cycling lanes on car traffic, and attempt to maximize utilization of said lanes with minimal impact on car travel times. Car usage is not considered by @mauttone2017bicycle, who develop an optimization framework that aims to minimize the total user cost of cycling on the network. The aggregate flow on links^[*link* refers to a road segment throughout this research] is obtained by using shortest paths to route existing cycling demand onto the road network, and the solution is a proposed set of links where cycling infrastructure should be added in order to minimize the overall travel cost of cyclists across the network. The cost of traversing a link is given as a function of its length and whether or not it has cycling infrastructure. The problem has also been solved by attempting to find the minimum cost of improving roadway links to meet a desired level of service (LOS) [@duthie2014optimization]. In this formulation, all origin-destination (OD)^[*Origin-Destination (OD)* data is normally obtained from a census. In this paper we use open access data from the UK census [@ONS2011flowdata], which contains aggregate statistics on number of commuters between Middle layer Super Output Area (MSOA) zones, by mode of travel. MSOAs have an average population of just over 8000 [@ofn2018population].] pairs need to be connected by roads that meet the desired LOS. A limitation of these problem formulations do not explicitly solve for continuity, which is dealt with using either a constraint specifying that each link with a bike lane should be connected to at least one destination [@mesbah2012bilevel], a constraint on maximum deviation from shortest paths [@duthie2014optimization], or a discontinuity penalty to prioritize connected road segments [@mauttone2017bicycle].

In this paper, continuity is analysed by looking at the connectivity of the network through the graph-theoretic concept of *connected components*. Other academics have taken this approach; @natera2019data study the existing cycling network in terms of its disconnected components and introduce two different algorithms to connect these components by their most critical links and, in doing so, measure the size of the growth of the largest connected component as a function of the kilometers of network added. The concept of connected components is also at the core of the methodology proposed by @olmos2020data. After routing the cycling demand onto the network links, they use percolation theory to filter out the links based on the aggregate flow^[*flow* is used throughout this research to refer to the cycling demand when it is routed onto the road network. The flow on any road segment is the cumulative demand on it, resulting from cyclists commuting between various OD pairs] passing through them, varying the flow threshold for filtering to identify the minimum flow at which the whole city is connected by a giant component. While these approaches deal with continuity better, they look at the network as a whole when attempting to improve it, and in doing so fail to account for equitable distribution of infrastructure.

## Ethical Underpinnings and Proposed Approach

All of these network-level methodologies are underpinned by ethical principles, even though these principles are not explicitly acknowledged by the authors. This is important since different ethical principles constitute different problem formulations and targets. Broadly speaking, transport appraisal can be based on either utilitarian or egalitarian principles. The former seeks to maximize the overall benefit, while the latter is concerned with a fair distribution of benefits [@jafino2020transport]. @nahmias2017integrating criticize the utilitarian approach that has been historically popular in the evaluation of transport investments, explaining how the maximization of overall benefit fails to account for the distribution of that benefit among communities or individuals. @lucas2016method explain how transport studies have traditionally looked at the bigger picture without studying the distribution of investments on the different parts of the study area, and go on to propose an egalitarian approach that ensures the disaggregation of transport policy benefits across the study area. @pereira2017distributive also emphasize the need for a more egalitarian approach to transport planning. They highlight accessibility as a cornerstone of distributive justice, and contend that policies should aim to distribute investments in a way that minimizes spatial variations in accessibility.

This research attempts to propose an egalitarian framework for cycling network design. This is done by identifying the different sub-networks that exist within the larger network, and ensuring they all get a fair share of investment. Trip patterns in a city are not uniformly distributed geographically, and *community finding* methods have been used to partition study areas into localized areas that experience a disproportionate number of trips within them. @akbarzadeh2018designing use a modularity maximization approach [@blondel2008fast] on taxi trip data to identify 7 different communities in the city of Isfahan, Iran. An optimization problem is then formulated to connect nodes within each community with cycling infrastructure, with the emphasis being on connectivity within the communities, not between them. @bao2017planning adopt a similar methodology, first identifying communities and then using a greedy network expansion algorithm to simultaneously add links to each community. The link with the highest benefit-cost ratio in each community is selected, and the network is grown by adding neighboring links to the solution until a budget limit is met. The benefit is the flow on the link, and each link is assigned a cost based on current road conditions.

<!-- EDIT! Add more to limitations -->
Our work builds on these *community finding* approaches by proposing a similar greedy network expansion algorithm for cycle network expansion within communities. We incorporate community finding methods for study area partitioning with weighted routing to avoid links that are stressful to cycle on. In doing so, we propose a methodology that accounts for motivators and deterrents to cycling. We propose three sub-methodologies that expand on some of the limitations of previous studies. These limitations include (a) bias inherent when basing network design solely on existing cycling demand, (b) proposing routes that may not correspond to studies on cyclist preference and government policies, and (3) insufficient consideration of the ethical principles underlying the analysis. Section \@ref(calculating-potential-cycling-demand) focuses on calculating potential cycling demand. Section \@ref(routing) focuses on routing the demand onto the road network while accounting for cyclist preferences and government priorities. Section \@ref(community-detection) outlines a methodology for partitioning the study area based on a community finding algorithm and routed cycling demand. Section \@ref(road-segment-prioritization) introduces the network expansion algorithms, and compares an approach grounded in ‘egalitarianism' to one grounded in ‘utilitarianism'.

@@

Relying only on existing cycling activity to prioritise cycling infrastructure can reinforce existing cycling patterns and ignores potential cycling demand that could be satisfied by a connected network. To avoid this issue, @duthie2014optimization choose to ignore existing demand completely, and focus on creating a network that connects the entire study area. @olmos2020data obtain the distance distribution of cyclists using a smartphone-based bicycle GPS data, and then use a rejection-sampling algorithm on the OD data of the study area to match the potential demand distribution to the distribution obtained from GPS data.

For our purposes, we use a logistic regression model to calculate potential cycling demand. The model was adapted from the Propensity to Cycle Tool (PCT), which estimates the proportion of trips ($\boldsymbol{C_{p}}$) for each OD pair that would cycle under different scenarios of change (we used the Government Target scenario, representing a nationwide target of doubling cycling by 2025) as a function of distance and hilliness [@lovelace2017propensity]. The logistic regression model used to calculate $\boldsymbol{C_{p}}$ has the following parameters:

@@

It should be noted that the calculations assume cycling in the traditional sense, i.e. the topography affects the decision to cycle. Recently there have been various micro-mobility solutions, including e-bikes, that allow commuters to traverse longer distances and hillier roads with less effort than traditional bicycles. While these modes would probably be associated with less geographical impedance, it is beyond the scope of this work to integrate those modes into the analysis. This is partially restricted by the lack of data on the proliferation of these modes, raising the point that perhaps the census data category of `Bicycle' is too vague, and should be further disaggregated to distinguish between traditional bicycles and other forms of micro-mobility.

@@

A limitation of the PCT approach outlined in the previous section is that it is reliant on external and therefore inflexible routing services. The next step in our approach is to route the potential cycling demand ($\boldsymbol{C_{p}}$) between all OD pairs onto the road network. We choose not to use the PCT approach for routing as it is reliant on external and therefore inflexible routing services.
<!-- This expands on the work of @mauttone2017bicycle, by going beyond simply favoring roads with existing cycling infrastructure to creating a hierarchy of road preference. -->

@@

\noindent A weighted distance $\boldsymbol{d_{w}}$ for each road segment is calculated as follows ^[The **dodgr** r package [@padgham2019dodgr] is used to route cycling demand onto the road network. The package uses the OpenStreetMaps (OSM) road network and allows the user to assign weights to roads based on their type. The routing is done based on weighted shortest paths, with the distance along each road segment being divided by a factor to obtain the weighted distance for routing. It is more intuitive to multiply when weighting a network, but the dodgr package divides by numbers between 0 and 1, which achieves the same result. For the sake of reproducibility, we stick to the convention used in the package.]:

@@

```{r flowsfacetunweighted, fig.cap="Flow Results Based on Unweighted Shortest Paths (Manchester)", out.width="90%", fig.show='hold', echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/flows_facet_unweighted_Manchester.png")
```

```{r flowsfacetweighted, fig.cap="Flow Results Based on Weighted Shortest Paths (Manchester)", out.width="90%", fig.show='hold', echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/flows_facet_weighted_Manchester.png")
```

@@

A major challenge facing ‘top-down’ planning approaches is how to incorporate egalitarian principles by fairly distributing investments in cycling infrastructure. One way of quantifying this is to split up the city into smaller geospatial areas and target equal investment in each of those areas, which could also help ensure that on-the-ground surveys are made by local stakeholders, an important component of the planning process  [@parkin2018designing]. Community detection offers us a way to delineate such a split; cyclists are limited in their commuting distance (see Figure \@ref(fig:cyclinghistmanc)), and so trip attractors are more likely to have a local catchment area of cyclists.

@@

In our case, the network is the city; the nodes are the population-weighted MSOA centroids and the links connecting each MSOA pair are weighted by the potential cycling demand between them. The Louvain method  [@blondel2008fast] is used to separate MSOAs into communities. Potential cycling demand is used since we assume that this is what the cycling demand will be once the cycling infrastructure is added. To assign road links to communities, the following steps are carried out:

<!-- 1. Create links between MSOA centroids and weigh these links by potential cycling demand between them.
2. Use Louvian method to determine optimal number of communities and assign each MSOA centroid to a community.
3. Assign each road link to the same community as the closest MSOA centroid to it. -->
```{r, eval = FALSE}
1. Create links between MSOA centroids and weigh these links by potential cycling demand 
   between them.
2. Use the Louvain method to determine the optimal number of communities and assign each MSOA 
   centroid to a community.
3. Assign each road link to the same community as the closest MSOA centroid to it.
```

@@

```{r growthtotal, fig.cap="Comparing Overall (Dashed) and Community Level Person-Km Satisfied (Manchester)", out.width="35%", fig.show='hold', echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/Growth_Results/growth_utilitarian_satisfied_km_both_flow_column.png",
  "data/Manchester/Plots/Growth_Results/growth_egalitarian_satisfied_km_both_flow_column.png"
))
```

@@

```{r growth3MapandBar, fig.cap="Results of Alg. 2 (Manchester)", out.width="35%", fig.show='hold', echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/Growth_Results/growth_egalitarian_priority_all_FLOW.png",
  "data/Manchester/Plots/Growth_Results/growth_egalitarian_investment_highways_flow.png"
))
```

@@

```{r componentsandGCC, fig.cap="Network Characteristics", out.width="35%", fig.show='hold', echo=FALSE}
knitr::include_graphics(c(
  "data/Manchester/Plots/Growth_Results/growth_util_egal_components_number_comparisonManchester.png",
  "data/Manchester/Plots/Growth_Results/growth_util_egal_components_gcc_comparisonManchester.png"
))
```

@@

While segregated, connected, and direct cycling infrastructure is key to achieving high levels of cycling, research has shown that it cannot exist in a vacuum. @wardman2007factors developed a mode choice model for the UK and their results showed that improved cycling infrastructure on its own only had modest impacts on mode shift, and even the unlikely scenario of all urban routes being serviced by segregated bike lanes was forecast to increase cycling mode share by only 3%. International research shows that cities that invest in more comprehensive cycling projects have a more significant increase in the number of cyclists as well as the cycling mode share [@parkin2018designing; @pucher2010infrastructure]. These cities do not just focus on infrastructure, but on general policies as well as restricting car use. Evaluation of policies in Denmark and Germany and the Netherlands has shown that their high cycling mode share is down to a broader set of policies that also include traffic calming, filtered permeability interventions, cycling rights of way, bike parking, integration with the public transport network, and making driving cars both expensive and inconvenient [@pucher2008making]. While these policies are outside the scope of this research, it is important to recognize their key role in bringing about an increase in levels of cycling.

@@

This paper provides a methodology for prioritizing road segments for investments in cycling infrastructure, with the output being a proposed cycling network for the city under study. 
The approach aims to respect both the needs of the users and the ambitions of stakeholders working at local/regional levels.
The results, primarily detailed route network maps based on current travel behaviour derived from OD data, can provide an evidence-based starting point for introducing a direct, connected, and low-stress network for any city by leveraging available open data. Given that the “most essential activity entailed in the design of cycle-friendly infrastructure is developing a cycle network” [@parkin2018designing], we believe that the methods have great potential to inform investment in cities such as Manchester where there is political will to invest in cycling long-term. A benefit of the approach is that it has relatively modest data requirements: only the road network (from OSM), its topography (from satellite imagery), and origin-destination data (from the national census in this case) are needed, opening up the possibility of deploying the approach in other cities^[The results are easily reproducible for all UK cities, and can also be reproduced for cities elsewhere given the availability of commuter data. For instructions on reproducing the results shown in this paper, see the **README** file in the R folder of the .zip file that accompanies this paper (to be released as open source software when the paper is published).].

@@

A core part of the methodology is determining where there is a high potential for cycling, and using this as a basis for recommendations on road space reallocation in order to unlock potential cycling demand. The framework provides a routing engine to study the effects of limiting access on different road types, finding that reducing cyclist flow on roads with high through-traffic results in acceptable increases in commuting distances. As a result, a hierarchy of road preference needs to be defined to route potential cycling demand. The framework provides different algorithms to determine where investments in cycling infrastructure should be prioritized. These algorithms, based on connectivity and egalitarian principles of resource distribution, ensured that whatever the level of investment, the resulting cycling network would improve connectivity of existing cycling infrastructure.

The methods can support the ongoing shift towards active transport, which is one of the priority areas of investments worldwide in transport planning. This necessary shift predates the pandemic and is driven by ambitions to improve the health and well-being of the citizens worldwide. The success of such a shift depends on the appeal of the cycling network for prospective cyclists, together with policies and interventions deterring the usage of the private car.
